Local Type Inference for Polarised System F with Existentials

### 2024-05-07 oopsla24/tex/06-extensions.tex
Edit extentions

In our system, all type applications are inferred implicitly: the algorithm
automatically instantiates the variables abstracted by $\forall$. 
The implicit type application can be added to the declarative system by the following rule:
$$\ottdruleDTTypeAppLabeled{}$$

However, this rule alone would cause ambiguity.
The declarative system does not fix the order of the quantifiers, 
which means that $[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ can be inferred 
as a type of $[[c]]$ interchangeably. But then explicit instantiation
$([[ c[iP] ]])$ would be ambiguous, as it is unclear whether $[[α⁺]]$ or
$[[β⁺]]$ should be instantiated with $[[iP]]$.

@@

The drawback of this approach is that it 
would cause the `leakage' of the internal algorithmic 
concept of type normalization into the `surface' declarative 
system.

@@

According to these rules, if two polymorphic ($[[∀]]$ or $[[∃]]$) types are
subtypes of each other, they must have the same top-level quantifiers. Moreover,
the equivalence on types (mutual subtyping) degenerates to \emph{equality} up to
alpha-conversion.

To accommodate these changes in the \emph{algorithm}, 
it suffices to 
    (i) replace the normalization procedure with identity: $[[nf(iN)]] \defeq [[iN]]$, $[[nf(iP)]] \defeq [[iP]]$,
        since the new equivalence classes consist of singletons;
    (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel}
        so that it requires the quantifiers to be equal (and performs alpha-conversion if necessary):
        $$\ottdruleLUBEExists{}$$ 
    (iii) replace the subtyping polymorphic rule \ruleref{\ottdruleAForallLabel} 
    by the following rule:
        $$\ottdruleAForallLR{}$$
    and
    (iv) update the existential rule \ruleref{\ottdruleAExistsLabel} symmetrically.

@@

\paragraph{Solution 3: Labeled Quantifiers}

A compromise solution that resolves the ambiguity of 
explicit the instantiation without fixing the order of
quantifiers is using \emph{labeled} quantifiers.
Let us discuss how to introduce them to the \emph{negative} part of the system,
since the positive subsystem is symmetric.

@@

The subtyping rule allows the quantifiers to have an arbitrary order. 
However, similarly to the elementary subtyping, 
each of the right-hand side quantifiers must have 
a matching left-hand side quantifier with the same label. These
matching quantifiers synchronously removed
(in other words, each variable is abstractly instantiated to itself), 
and the remaining quantifiers are instantiated as usual by
a substitution $[[σ]]$.
$$ \ottdruleDOneLblForallLRLabeled{} $$

@@

\subsection{Weakening of the Subtyping Invariant}
\label{sec:weakening-invariant}

@@

This way, the algorithm can be extended to support the \emph{covariant} up-shift
subtyping. This, for example, enriches the relation with such subtypes as $[[· ⊢
∀α⁺.α⁺→↑α⁺ ≤ ↓↑Int → ↑∃β⁻.↓β⁻]]$, which does not hold in the original system.
Moreover, this extension also increases the expressiveness of the \emph{type
inference}. Namely, when inferring a type of an annotated let binding 
$[[let x:iP = v(args); c]]$, we require $[[↑iP]]$ to be a supertype of 
$[[↑iQ]]$---the resulting type of the application $[[v]]([[args]])$. 
Now, it is equivalent to the requirement that $[[iP]]$ is a supertype of $[[iQ]]$
(without the up-shift), which is more permissive.

In addition, the enhancement of the unification algorithm described above
makes possible another improvement to the system---bounded quantification.

@@

For brevity, we do not discuss in detail the combination of bounded and
unbounded quantification in one system. However, this update is straightforward:
the unbounded quantifiers are treated as the weakest bound, which is trivially
satisfied. Analogously, it is possible to extend the system with \emph{lower}
bound $[[∀]]$-quantifiers: $[[∀(pa ≥* iP). iN]]$. 
Then the generated constraints will change to $[[βî⁺ :≥ iQi]]$, which 
is also covered by the resolution procedure.
However, we do not consider bounded \emph{existential} quantification,
because in this case, the constraint resolution would require us to 
find the \emph{greatest lower bound} of two negative types, which is not
well-defined (see \cref{sec:constraint-merge}).

@@

The algorithm we provide requires that all lambda functions are annotated. 
This restriction significantly simplifies the type inference by making the 
terms uniquely define their types. However, it leads to certain redundancies
in typing, in particular, the type of a lambda expression cannot be inferred 
from the context it is used in: $[[((λ x . return x) : (Int → ↑Int))]]$
is not allowed.

@@

    For brevity, we omit the details of the bidirectional algorithm.
    However, let us discuss one especially important rule---arrow application inference. 
    An intuitive way to bidirectionalize this rule is the following:
    $$ \ottdruleATArrowAppBidirLabeled{} $$\\
    % \paragraph{Reaching Undecidability}
    However, this rule is too allowing. 
    The judgement $[[Γ ; Φ ⊨ v ⇐ uQ ⫤ SC1]]$
    requires us to check a term against an \emph{algorithmic} 
    type $[[uQ]]$. Further, through the
    lambda function checking \ruleref{\ottdruleDTtLamChkLabel}
    the algorithmic types infiltrate the type context
    and then through variable inference and subsumption, 
    \emph{both} sides of subtyping. 
    This way, \ruleref{\ottdruleATArrowAppBidirLabel}
    compromises the important invariant
    of the subtyping algorithm: 
    now the same algorithmic variable can occur on \emph{both}
     sides of the subtyping relation.

    We believe that the relaxation of this invariant brings the 
    constraint resolution too close to the second-order pattern 
    unification, which is undecidable \cite{goldfarb81:undecidability}.
    In particular, the constraint $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable with 
    $[[α̂⁺]] = [[∃β⁻.↓β⁻]]$, 
    since by \ruleref{\ottdruleDOneExistsLabel}, 
    the existential $[[β⁻]]$ can be impredicatively instantiated to $[[↑∃β⁻.↓β⁻]]$.

    The constraints such as $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ are not merely hypothetical. 
    It occurs, for example, when we infer the type of the following application:
    $$[[· ; · ⊢ ∀α⁺ . ↓(α⁺ → ↑Int) → ↓↑α⁺ → ↑Int ● {λx.λy.let y' = x(y); return y'} ⇒> ↑Int]]$$
    For $[[x]]$ to be applicable to $[[y]]$,
    the type of $[[y]]$---$[[↓↑α⁺]]$ must be a subtype of
    $[[α⁺]]$---the type expected by $[[x]]$. This way, 
    this inference is possible if and only if
    $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable.
    Using a similar scheme, we can construct different examples
    whose resolution significantly relies on second-order pattern unification.
    Thus, we leave the resolution of this type of constraints beyond the scope of this work.

>>>>

In our system, all type applications are inferred implicitly: the algorithm
automatically instantiates the variables abstracted by $\forall$. 
The \emph{explicit} type application can be added to the declarative system by the following rule:
$$\ottdruleDTTypeAppLabeled{}$$

However, this rule alone would cause ambiguity. The declarative system does not
fix the order of the quantifiers, which means that $[[∀α⁺.∀β⁺.iN]]$ and
$[[∀β⁺.∀α⁺.iN]]$ can be inferred as a type of $[[c]]$ interchangeably. But that
would make explicit instantiation $[[ c[iP] ]]$ ambiguous, as it is unclear
whether $[[α⁺]]$ or $[[β⁺]]$ should be instantiated with $[[iP]]$.

@@

The drawback of this approach is that it adds another point where the internal
algorithmic concept---normalization---is exposed to the `surface' declarative
system.

@@

According to these rules, if two quantified ($[[∀]]$ or $[[∃]]$) types are
subtypes of each other, they must have the same top-level quantifiers. Moreover,
the equivalence on types (mutual subtyping) degenerates to \emph{equality} up to
alpha-conversion.

To accommodate these changes in the \emph{algorithm}, 
it suffices to 
    (i) replace the normalization procedure with identity: $[[nf(iN)]] \defeq [[iN]]$, $[[nf(iP)]] \defeq [[iP]]$,
        since the new equivalence classes are singletons;
    (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel}
        so that it requires the quantifiers to be equal (and performs alpha-conversion if necessary):
        $$\ottdruleLUBEExists{}$$ 
    (iii) replace the subtyping polymorphic rule \ruleref{\ottdruleAForallLabel} 
    by the following rule:
        $$\ottdruleAForallLR{}$$
    and
    (iv) update the existential rule \ruleref{\ottdruleAExistsLabel} symmetrically.

@@

\paragraph{Solution 3: labeled quantifiers}

A compromise solution that resolves the ambiguity of explicit instantiation
without fixing the order of quantifiers is by using \emph{labeled} quantifiers.
Let us discuss how to introduce them to the \emph{negative} part of the system,
since the positive subsystem is symmetric.

@@

The subtyping rule permits an arbitrary order of the quantifiers.
However, similarly to the elementary subtyping, 
each of the right-hand side quantifiers must have 
a matching left-hand side quantifier with the same label. These
matching quantifiers synchronously removed
(in other words, each variable is abstractly instantiated to itself), 
and the remaining quantifiers are instantiated as usual by
a substitution $[[σ]]$.
$$ \ottdruleDOneLblForallLRLabeled{} $$

@@

\subsection{Weakening of the subtyping invariants}
\label{sec:weakening-invariant}

@@

This way, the algorithm can be extended to support the \emph{covariant} up-shift
subtyping. This, for example, enriches the relation with such subtypes as $[[· ⊢
∀α⁺.α⁺→↑α⁺ ≤ ↓↑Int → ↑∃β⁻.↓β⁻]]$, which does not hold in the original system.
Moreover, this extension also increases the expressiveness of the \emph{type
inference}. Namely, when inferring a type of an annotated let binding $[[let
x:iP = v(args); c]]$, we require $[[↑iP]]$ to be a supertype of $[[↑iQ]]$---the
resulting type of the application $[[v]]([[args]])$. In the relaxed system, it
can be replaced with the requirement that $[[iP]]$ is a supertype of $[[iQ]]$
(without the up-shift), which is more permissive.

In addition, the refinement of the unification algorithm described above
makes possible another improvement to the system---bounded quantification.

@@

For brevity, we do not discuss in detail the combination of bounded and
unbounded quantification in one system. We believe this update is straightforward:
the unbounded quantifiers are treated as the weakest constraints, which are trivially
satisfied. Analogously, it is possible to extend the system with \emph{lower}
bound $[[∀]]$-quantifiers: $[[∀(pa ≥* iP). iN]]$. 
Then the generated constraints will change to $[[βî⁺ :≥ iQi]]$, which 
is also covered by the resolution procedure.
However, we do not consider bounded \emph{existential} quantification,
because in this case, the constraint resolution would require us to 
find the \emph{greatest lower bound} of two negative types, which is not
well-defined (see \cref{sec:constraint-merge}).

@@

The algorithm we provide requires that all lambda functions are annotated. 
This restriction significantly simplifies the type inference by making the 
terms uniquely define their types. However, it leads to certain redundancies
in typing, in particular, the type of a lambda expression cannot be inferred 
from the context it is used in: the annotated $[[((λ x . return x) : (Int → ↑Int))]]$
does not infer $[[Int → ↑Int]]$.

@@

    For brevity, we omit the details of the bidirectional algorithm.
    Let us discuss one especially important rule---arrow application inference. 
    An intuitive way to bidirectionalize this rule is the following:
    $$ \ottdruleATArrowAppBidirLabeled{} $$
    % \paragraph{Reaching Undecidability}
    However, this rule is too permissive. 
    The judgement $[[Γ ; Φ ⊨ v ⇐ uQ ⫤ SC1]]$
    requires us to check a term against an \emph{algorithmic} 
    type $[[uQ]]$. Further, through the
    lambda function checking \ruleref{\ottdruleDTtLamChkLabel}
    the algorithmic types infiltrate the type context
    and then through variable inference and subsumption, 
    \emph{both} sides of subtyping. 
    This way, \ruleref{\ottdruleATArrowAppBidirLabel}
    compromises the important invariant
    of the subtyping algorithm: 
    now the same algorithmic variable can occur on \emph{both}
     sides of the subtyping relation.

    We believe that the relaxation of this invariant brings the 
    constraint resolution too close to second-order pattern 
    unification, which is undecidable \cite{goldfarb81:undecidability}.
    In particular, the constraint $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable with 
    $[[α̂⁺]] = [[∃β⁻.↓β⁻]]$, 
    since by \ruleref{\ottdruleDOneExistsLabel}, 
    the existential $[[β⁻]]$ can be impredicatively instantiated to $[[↑∃β⁻.↓β⁻]]$.

    Constraints such as $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ are not merely hypothetical. 
    It occurs, for example, when we infer the type of the following application:
    $$[[· ; · ⊢ ∀α⁺ . ↓(α⁺ → ↑Int) → ↓↑α⁺ → ↑Int ● {λx.λy.let y' = x(y); return y'} ⇒> ↑Int]]$$
    For $[[x]]$ to be applicable to $[[y]]$,
    the type of $[[y]]$---$[[↓↑α⁺]]$ must be a subtype of
    $[[α⁺]]$---the type expected by $[[x]]$. This way, 
    this inference is possible if and only if
    $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable.
    Using a similar scheme, we can construct different examples
    whose resolution significantly relies on second-order pattern unification.
    Thus, we leave the resolution of this type of constraints beyond the scope of this work.
